Dynamics of innovation in stigmergic collectives

### 2023-04-27 LaTeX/draft.tex
MoreChngs

\def\grad{\nabla}
\def\at{\hat{a}}
\def\Ai{\mathcal{A}}
\def\H{\mathcal{H}}
\def\J{\mathcal{J}}
\def\R{\mathbb{R}}
\def\d{\text{d}}
\def\e{\text{e}}
\def\r{\mathbf{r}}
\def\x{\mathbf{x}}
\def\xst{\mathbf{x}^*}
\def\u{\mathbf{u}}
\def\a{\mathbf{a}}
\def\T{\mathsf{T}}
\def\R{\mathbb{R}}
\def\E{\mathbb{E}}
\def\S{\mathcal{S}}
\def\pol{\pi_\theta}
\def\qsa{Q (s, a)}
\def\paa{\pi(a, s)}
\def\H{\sigma}
\def\grad{\nabla}
\def\th{\hat{\mathbf{t}}}
\def\ph{\hat{\mathbf{p}}}
\def\nh{\hat{\mathbf{n}}}
\def\yd{\dot{y}}
\def\ite{\text{ite}}
\def\P{\mathcal{P}}
\def\D{\mathcal{D}}
\def\Z{\mathcal{Z}}
\def\ah{\hat{a}}
\def\theta{\vartheta}
\def\H{\mathcal{H}}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

%\pacs{Valid PACS appear here}
\maketitle
\section{Problem setting}
Let us consider a trail of pheromone $\x^*(s)$ representing the position in 2-D coordinates and parameterized by the
arc-length $s$. The concentration of pheromone is considered continuous along this curve and is given by $c(\x) = c_o \delta(\x-\x^*(s))$.
This field is of course steady and its dynamics will be considered later in the article. The state of the agent/ant in
the environment is is represented by its position $\r^t=\{ r_x^t, r_y^t \}$ and the action it can take is to choose the
direction of motion i.e., its orientation $\ph^t = \{\cos(\theta^t), \sin(\theta^t)\}$, where $[\bullet]^t$ denotes the
value of the variable at time-instant $t$. The agent can choose the action based on the measurements at each instant of time.
In this article we are interested in understanding the dynamics of agents that need to innovate in order to find new
trails over which the collective travels. In order to understand the dynamics, we assume that the agent has already
traveled along the trajectory $\x^*(s)$ once before.
In this article we are interested in understanding how and when an agent can innovate from its innate behavior to follow
pheromone trail to explore the environment and find novel solutions to the navigation problem.

\subsection*{Trail following behavior for semi-circular trail}
Social insects such as ants communicate either through pheromone or by antennation.
As stigmergy is one of the fundamental mechanisms by which the collective coordinates
and recruits new workers, we assume that the agent can follow the trail $\x^*(s)$.
This intrinsic behavior is denoted by the policy $\pi_{\rm int}(\r)$ which provides the
appropriate angle $\theta^t$ required to follow the trail when the agent is at location
$\r^t$ after the measurment that the agent makes. In order to consider a specific trail
of the initial pheromone trail, let us assume that the trail is a simple semi-circle with
functional form $\xst(s) = \{ a \cos(s/a), a \sin(s/a) \}$. For this particular case,
the policy can be explicitly written as
\begin{align}
    \text{Measurement: } \phi^{t} = & \ \tan^{-1} \bigg( \frac{r_y^t}{r_x^t} \bigg) + \zeta^t, \\
    \text{Action: }\theta^{t} =& \ \frac{\pi}{2} + \phi^{t}, \\
    \text{State update: } \r^{t+1} =& \ \r^t + l \ph^{t}.
\end{align}
where $\zeta^t$ is the fluctuation/error in the measurements that the agent makes, $l$ is the step-length
of the agent. In the limit when the step-length is small, we can write the continuous version
of the motion as $\dot{\r} = v_o \ph$. Without loss of generality, we can write this as
$\d\r/\d s = \ph$ where $s = v_o t$ and we assume that the distribution of the orientation of the agent is distributed as though
it is a worm-like chain polymer whose orientation distribution is given by
\[
    \P[\theta(s)] = \frac{1}{\mathcal{Z}}\exp{(-\beta E_b)}
\]
where the bending energy $E_b[\theta(s)] = B/2 \int_0^{s_c} (\theta'(s)-\kappa_o)^2 \ \d s$, the partition
function $\mathcal{Z} = \int \mathcal{D}[\theta(s)] \exp\{ {-\beta E_b[\theta(s)]} \}$ and the bending
stiffness $B$ captures the constraint for the agent to move forward, $\beta$ is the effective
temperature capturing the fluctuations in the trajectory of agent. We also have
boundary conditions $\theta(0) = \pi/2$ and $\theta'(0) = \kappa_o = a^{-1}$. Here $s_c$ is the coordinate
along the semi-circle at which the agent bifurcates towards the target location A using a new trajectory.
Our motivation is to identify this location using the Bellman equation which will be discussed later.
Now in order to evaluate the quality of the intrinsic policy we introduce the notion of value function
of this policy $\pi_{\rm int}(\r)$,denoted by $v_{\rm int}(\r^t)$. The value function provides the cummulative
future reward that the agent will acquire when following the policy $\pi_{\rm int}(\r)$. This can be written as
\[
    v_{\rm int}(\r) = \E_\pi \bigg[ \sum_{k=0}^\infty \gamma^k R_{t+k+1} \bigg| \r_t = \r \bigg].
\]
We can write this in path-integral form as
\begin{align}
    v_{\rm int}(\r) =& \ \sum_{k=t}^T \int \D[\theta(s)] \P[\theta(s)] \gamma^{k-t} R_{k+1}, \\
    \approx& \ \int_{s}^{s_c} \int \D[\theta(s)] \P[\theta(s)] R(s) \ \d s
\end{align}
The reward itself is a function of the agent's azimuthal direction, $\phi(s) = \tan^{-1}(y(s)/x(s))$.
Assuming that the reward has the function form $R(\phi) = \exp(-\phi/\phi^*)$, the leading order form of 
the value function for this policy is
\[
    v_{\rm int}(\r) = \frac{\phi^*}{\kappa_o} [ \e^{-\kappa_os/\phi^*} - \e^{-\kappa_o s_c/\phi^*}].
\]

@@

\subsection*{Search strategy post bifurcation}
After the agent leaves the trail trajectory at $\r(s_c)$, the agent needs to reach the target A.

>>>>

\def\grad{\nabla}
\def\at{\hat{a}}
\def\Ai{\mathcal{A}}
\def\H{\mathcal{H}}
\def\J{\mathcal{J}}
\def\R{\mathbb{R}}
\def\d{\text{d}}
\def\e{\text{e}}
\def\r{\mathbf{r}}
\def\x{\mathbf{x}}
\def\xst{\mathbf{x}^*}
\def\u{\mathbf{u}}
\def\a{\mathbf{a}}
\def\T{\mathsf{T}}
\def\R{\mathbb{R}}
\def\E{\mathbb{E}}
\def\S{\mathcal{S}}
\def\pol{\pi_\theta}
\def\qsa{Q (s, a)}
\def\paa{\pi(a, s)}
\def\H{\sigma}
\def\grad{\nabla}
\def\th{\hat{\mathbf{t}}}
\def\ph{\hat{\mathbf{p}}}
\def\nh{\hat{\mathbf{n}}}
\def\yd{\dot{y}}
\def\ite{\text{ite}}
\def\P{\mathcal{P}}
\def\D{\mathcal{D}}
\def\Z{\mathcal{Z}}
\def\ah{\hat{a}}
\def\theta{\vartheta}
\def\H{\mathcal{H}}
\DeclareMathOperator*{\argmax}{arg\,max}
\def\OU{\rm OU}
\def\intP{\rm int}

@@

%\pacs{Valid PACS appear here}
\maketitle
\section{Problem setting}
Let us consider a trail of pheromone $\x^*(s)$ representing the position in 2-D coordinates and parameterized by the
arc-length $s$. The concentration of pheromone is considered continuous along this curve and is given by $c(\x) = c_o \delta(\x-\x^*(s))$.
This field is of course steady and its dynamics will be considered later in the article. The state of the agent/ant in
the environment is is represented by its position $\r^t=\{ r_x^t, r_y^t \}$ and the action it can take is to choose the
direction of motion i.e., its orientation $\ph^t = \{\cos(\theta^t), \sin(\theta^t)\}$, where $[\bullet]^t$ denotes the
value of the variable at time-instant $t$. The agent can choose the action based on the measurements at each instant of time.
In this article we are interested in understanding the dynamics of agents that need to innovate in order to find new
trails over which the collective travels.

\subsection*{Trail following behavior for semi-circular trail}
Social insects such as ants communicate either through pheromone or by antennation.
As stigmergy is one of the fundamental mechanisms by which the collective coordinates
and recruits new workers, we assume that the agent can follow the trail $\x^*(s)$.
In this article we are interested in understanding how and when an
agent can innovate from its innate behavior to follow pheromone trails to explore the environment and find novel solutions
to the navigation problem. In order to understand the dynamics of innovation, we assume that the agent has already
traversed the trajectory $\x^*(s)$ before.
This intrinsic behavior is denoted by the policy $\pi_{\rm int}(\r)$ which provides the
appropriate angle $\theta^t$ required to follow the trail when the agent is at location
$\r^t$ after the measurment that the agent makes. In order to consider a specific trail
of the initial pheromone trail, let us assume that the trail is a simple semi-circle with
functional form $\xst(s) = \{ a \cos(s/a), a \sin(s/a) \}$. For this particular case,
the policy can be explicitly written as
\begin{align}
    \text{Measurement: } \phi^{t} = & \ \tan^{-1} \bigg( \frac{r_y^t}{r_x^t} \bigg) + \zeta^t, \\
    \text{Action: }\theta^{t} =& \ \frac{\pi}{2} + \phi^{t}, \\
    \text{State update: } \r^{t+1} =& \ \r^t + l \ph^{t}.
\end{align}
where $\zeta^t$ is the fluctuation/error in the measurements that the agent makes, $l$ is the step-length
of the agent. In the limit when the step-length is small, we can write the continuous version
of the motion as $\dot{\r} = v_o \ph$. Without loss of generality, we can write this as
$\d\r/\d s = \ph$ where $s = v_o t$ and we assume that the distribution of the orientation of the agent is distributed as though
it is a worm-like chain polymer whose orientation distribution is given by
\[
    \P[\theta(s)] = \frac{1}{\mathcal{Z}}\exp{(-\beta E_b)}
\]
where the bending energy $E_b[\theta(s)] = B/2 \int_0^{s_c} (\theta'(s)-\kappa_o)^2 \ \d s$, the partition
function $\mathcal{Z} = \int \mathcal{D}[\theta(s)] \exp\{ {-\beta E_b[\theta(s)]} \}$ and the bending
stiffness $B$ captures the constraint for the agent to move forward, $\beta$ is the effective
temperature capturing the fluctuations in the trajectory of agent. We also have
boundary conditions $\theta(0) = \pi/2$ and $\theta'(0) = \kappa_o = a^{-1}$. Here $s_c$ is the coordinate
along the semi-circle at which the agent bifurcates towards the target location A using a new trajectory.
Our motivation is to identify this location using the Bellman equation which will be discussed later.
Now in order to evaluate the quality of the intrinsic policy we introduce the notion of value function
of this policy $\pi_{\rm int}(\r)$,denoted by $v_{\rm int}(\r^t)$. The value function provides the cummulative
future reward that the agent will acquire when following the policy $\pi_{\rm int}(\r)$. This can be written as
\[
    v_{\rm int}(\r) = \E_\pi \bigg[ \sum_{k=0}^\infty \gamma^k R_{t+k+1} \bigg| \r_t = \r \bigg].
\]
We can write this in path-integral form as
\begin{align}
    v_{\rm int}(\r) =& \ \sum_{k=t}^T \int \D[\theta(s)] \P[\theta(s)] \gamma^{k-t} R_{k+1}, \\
    \approx& \ \int_{s}^{s_c} \int \D[\theta(s)] \P[\theta(s)] R(s) \ \d s
\end{align}
The reward itself is a function of the agent's azimuthal direction, $\phi(s) = \tan^{-1}(y(s)/x(s))$.
Assuming that the reward has the function form $R(\phi) = \exp(-\phi/\phi^*)$, the leading order form of 
the value function for this policy is
\[
    v_{\rm int}(\r) = \frac{\phi^*}{\kappa_o} [ \e^{-\kappa_os/\phi^*} - \e^{-\kappa_o s_c/\phi^*}].
\]

@@

\subsection*{Search strategy post bifurcation}
After the agent leaves the trail trajectory at $\r(s_c)$, the agent needs to reach the target A.
Though the agent can take a variety of exploration strategies, we assume that the policy resembles
an Ornstein-Uhlenbeck (OU) process. This explorative policy $\pi_{\OU}(\r)$ can be written through $\P(\theta, t)$
the probability that the agent will take a particular angle $\theta$ at time $t$ or distance $s$. However since the orientation of the
agent is dependent on $\r(s_c)$, the angle between agent location $\r(s_c)$ to target A can be denoted
as $\phi$ and $\theta=\phi+\varphi$ where $\varphi$ is the fluctuation around this orientation that follows
an OU-process. The probability $\P(\varphi, s)$ can then be written as
\[
    \frac{\partial \P}{\partial s} = \frac{\nu}{v_o} \frac{\partial(\varphi \P)}{\partial \varphi }
    + \frac{D}{v_o} \frac{\partial^2 \P}{\partial \varphi^2}
\]
where $\nu$ is the time-scale associated with relaxation of the agent's orientation to the target $\phi$
and $D$ represents the fluctuation in orientation, which we will soon assume to be large when $\r(s_c)$
is farther from the target A. We assume that the agent reaching a distance $\sigma$ from the end-point A can reach the
target and thus the reward is a function of the propagation of variance in agent's position towards the
target. Instead of solving the partial differential equation for $\P(\varphi, s)$, we can evaluate the
evolution of the variance of agent's position $\r(s)$ as it marches close to the target.

Assuming that the agent reaching a distance $\sigma$ from the end-point A can reach the
target, the probability that an agent will reach the target is given by $\P_{\rm OU}(||\r^{n\Delta t} - \r^*|| \leq \sigma)
= \sigma/\psi(t^*)$. We can now calculate the value function, $v_{\rm OU}(\r, \theta)$ using the formula
\[
    v_{\rm OU}(\r, \theta) = \int_{0}^{t^*} \int \D[\theta(t)] \P_{\rm OU}[\theta(t)] R(t) \ \d t.
\]
Let us assume that the reward function for the OU process is $R(t) = \nu[1 - \Theta(||\r(t^*) - \r^*|| - \sigma)]$.
From this we see that the value of $v_{\rm OU}(\r^c, \theta^c) = \sigma \nu/\psi(t^*)$.

%\sectionfont{\fontsize{15}{15}\selectfont}
%\subsectionfont{\fontsize{15}{15}\selectfont}
%\subsubsectionfont{\fontsize{15}{15}\selectfont}
\large
\widetext
\clearpage
\onecolumngrid
\begin{center}
\textbf{\large Supplemental Information for ``Branching in soft gels''}\\[.2cm]
S  Ganga  Prasath$^{1, *}$, Aditi Chakrabarti$^{1, *}$,  and  L  Mahadevan$^{1, 2, 3, 4}$\\[.1cm]
{\small \itshape ${}^1$School of Engineering and Applied Sciences, Harvard University, Cambridge MA 02138.\\
${}^2$Department of Physics, Harvard University, Cambridge MA 02138.\\
${}^3$Center  for  Brain  Science,  Harvard  University,  Cambridge,  MA  02138.\\
${}^4$Department of Organismic and Evolutionary Biology, Harvard University, Cambridge 02138
}
\end{center}
%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\makeatletter
% \renewcommand{\theequation}{S\arabic{equation}}
% \renewcommand{\thefigure}{S\arabic{figure}}
% \renewcommand{\bibnumfmt}[1]{[S#1]}
%\renewcommand{\citenumfont}[1]{S#1}
%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\linespread{1.5}
% \setstretch{1.5}

%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
%\renewcommand{\bibnumfmt}[1]{[S#1]}
%\renewcommand{\citenumfont}[1]{S#1}
%\renewcommand{\thesection}[1]{S#1\arabic{section}}
\renewcommand{\thetable}{S\arabic{table}}%
\renewcommand{\thesection}{S\arabic{section}}
%\renewcommand{\thesubsection}{SS\arabic{subsection}}
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
%\renewcommand{\thesubsection}{SS\arabic{subsection}{}}
\renewcommand{\thesubsection}{\roman{subsection}{}}
%\renewcommand{\thesubsubsection}{(\roman{subsubsection}{})}
